Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-11-14 paper/paper.tex
appendix wip

\paragraph{$I\beta$}
Given
$\Delta'_{1},\cdot,\Delta'_{2} \vdash e' : C$, the desired
$\beta$ law is
\[
  \semg{\letin {()} {()} {e'}} = \semg{e}
\]
\begin{proof}
\begin{align*}
  \semg{\letin {()} {()} {e'}}
  & = \semg{e'} \circ m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ (id \otimes \semg{()}) \otimes id \circ m_{\Delta_{1},\cdot} \otimes id \circ m_{(\Delta'_{1},\cdot),\Delta'_{2}} \\
  & = \semg{e'} \circ m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ (id \otimes id) \otimes id \circ m_{\Delta_{1},\cdot} \otimes id \circ m_{\Delta'_{1},\Delta'_{2}} \\
  & = \semg{e'} \circ m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ \rho^{-1} \otimes id \circ m_{\Delta'_{1},\Delta'_{2}} \\
  & = \semg{e'}
\end{align*}
\end{proof}

\paragraph{$I\eta$}
Similarly, for $\Delta_{1}, a : A, \Delta_{3} \vdash e' : C$ and
$\Delta_{2} \vdash e : I$ the desired $\eta$ law is
\[
\semg{\letin {()} {e} {e'[()/a]}} = \semg{e'[e/a]}
\]
However, through application of \cref{lem:subst} it suffices to handle the case
where $e$ is a variable $a'$. That is,
\begin{align*}
\letin {()} {e} {e'[()/a]} &= (\letin {()} {a'} {e'[()/a]})[e/a'] \\
e'[e/a] &= e'[a'/a][e/a]
\end{align*}
so without loss of generality we may take may take $e$ to be variable $a'$.

\begin{proof}
\begin{align*}
  \semg{\letin {()} {a'} {e'[()/a]}}
  & = \semg{e'} \circ m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ (id \otimes \semg{a'}) \otimes id \circ m_{\Delta_{1},\cdot} \otimes id \circ m_{\Delta'_{1},\Delta'_{2}} \\
  & = \semg{e'} \circ m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ (id \otimes \lambda) \otimes id \circ m_{\Delta_{1},\cdot} \otimes id \circ m_{\Delta'_{1},\Delta'_{2}} \\
  & = \semg{e'}
\end{align*}

@@

Further by \cref{lem:subst},
\begin{align*}
  \semg{e'[a'/a]}
  &= \semg{e'} \circ \semg{a'/a} \\
  &= \semg{e'}
\end{align*}
Again by the coherence theorem, $\semg{a'a} = id$. Thus, $\eta$ law for $I$
holds in the denotational semantics.
\end{proof}

@@

\begin{tikzcd}
  \semg{\Delta,\Delta'} \arrow[r, "m_{\Delta,\Delta'}"] &
  \semg{\Delta} \otimes \semg{\Delta'} \arrow[r, "\semg{e_1} \otimes \semg{e_2}"] &
  \semg{A} \otimes \semg{G}
\end{tikzcd}

@@

\begin{tikzcd}
  \semg{\Delta'_1,\Delta,\Delta'_2} \arrow[r , "m_{(\Delta'_1,\Delta),\Delta'_2}"] &
  \semg{\Delta'_1,\Delta} \otimes \semg{\Delta'_2}
    \arrow[r , "m_{\Delta'_1,\Delta} \otimes id"] &
  \left(\semg{\Delta'_1} \otimes \semg{\Delta} \right) \otimes \semg{\Delta'_2}
    \arrow[d , "(id \otimes \semg{e}) \otimes id"] \\
  \semg{c} & \arrow[l , "\semg{f}"]
  \left( \left( \semg{\Delta'_1} \otimes \semg{A} \right) \otimes \semg{B}\right) \otimes \semg{\Delta'_2}
  & \left(\semg{\Delta'_1} \otimes \left( \semg{A} \otimes \semg{B}\right) \right) \otimes \semg{\Delta'_2}
    \arrow[l , "\alpha \otimes id"]
\end{tikzcd}

\paragraph{$\otimes\beta$}

\paragraph{$\otimes\eta$}

% linear function
In the semantics $\semg{- \otimes A}$ and $\semg{A \lto -}$ are left adjoint, so
for all $A$ and $B$ we have the natural isomorphism
$\phi : Hom(\semg{\Delta} \otimes \semg{A}, \semg{B}) \to Hom(\semg{\Delta}, \semg{A \lto B})$.
We use this adjunction to provide the denotation $\semg{\lamblto a e} : \semg{\Delta} \to \semg{A \lto B}$.

@@

The denotation for application of a $\lto$-function,
$\semg{\applto {e'} {e}} : \semg{\Delta,\Delta'} \to \semg{B}$ defined by the diagram

\begin{tikzcd}
  \semg{\Delta,\Delta'} \arrow[r , "m_{\Delta,\Delta'}"] &
  \semg{\Delta} \otimes \semg{\Delta'} \arrow[r , "id \otimes \semg{e}"] &
  \semg{\Delta} \otimes \semg{A} \arrow[r , "\phi^{-1}\left( \semg{e'} \right)"]
  &
  \semg{B}
\end{tikzcd}

$\semg{\lambtol a e} : \semg{\Delta} \to \semg{B \tol A}$ defined using the
natural isomorphism
$\psi : Hom(\semg{A} \otimes \semg{\Delta}, \semg{B}) \to Hom(\semg{\Delta}, \semg{B \tol A})$
acting on the following diagram

\begin{tikzcd}
  \semg{A} \otimes \semg{\Delta} \arrow[r , "\lambda^{-1} \otimes id"] &
  \semg{a : A} \otimes \semg{\Delta} \arrow[r , "m^{-1}_{a : A , \Delta}"] &
  \semg{a : A , \Delta} \arrow[r , "\semg{e}"] &
  \semg{B}
\end{tikzcd}

The application of a $\tol$-function,
$\semg{\apptol {e} {e'}} : \semg{\Delta',\Delta} \to \semg{B}$ defined by the diagram

\begin{tikzcd}
  \semg{\Delta',\Delta} \arrow[r , "m_{\Delta',\Delta}"] &
  \semg{\Delta'} \otimes \semg{\Delta} \arrow[r , "\semg{e'} \otimes id"] &
  \semg{A} \otimes \semg{\Delta} \arrow[r , "\phi^{-1}\left( \semg{e} \right)"] &
  \semg{B}
\end{tikzcd}

>>>>

\paragraph{$I\beta$}
Given
$\Delta'_{1},\cdot,\Delta'_{2} \vdash e' : C$, the desired
$\beta$ law is
\[
  \semg{\letin {()} {()} {e'}} = \semg{e}
\]
\begin{proof}
\begin{align*}
  \semg{\letin {()} {()} {e'}}
  & = \semg{e'} \circ m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ (id \otimes \semg{()}) \otimes id \circ m_{\Delta_{1},\cdot} \otimes id \circ m_{(\Delta'_{1},\cdot),\Delta'_{2}} \\
  & = \semg{e'} \circ m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ (id \otimes id) \otimes id \circ m_{\Delta_{1},\cdot} \otimes id \circ m_{\Delta'_{1},\Delta'_{2}} \\
  & = \semg{e'} \circ m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ \rho^{-1} \otimes id \circ m_{\Delta'_{1},\Delta'_{2}} \\
  & = \semg{e'} \tag{coherence}
\end{align*}
\end{proof}

\paragraph{$I\eta$}
Similarly, for $\Delta_{1}, a : A, \Delta_{3} \vdash e' : C$ and
$\Delta_{2} \vdash e : I$ the desired $\eta$ law is
\[
\semg{\letin {()} {e} {e'[()/a]}} = \semg{e'[e/a]}
\]
However, through application of \cref{lem:subst} it suffices to handle the case
where $e$ is a variable $a'$. That is,
\begin{align*}
\letin {()} {e} {e'[()/a]} &= (\letin {()} {a'} {e'[()/a]})[e/a'] \\
e'[e/a] &= e'[a'/a][e/a]
\end{align*}
so without loss of generality we may take may take $e$ to be variable $a'$. We
will additionally use this style of argumentation when necessary throughout this section.

\begin{proof}
\begin{align*}
  \semg{\letin {()} {a'} {e'[()/a]}}
  & = \semg{e'} \circ m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ (id \otimes \semg{a'}) \otimes id \circ m_{\Delta_{1},\cdot} \otimes id \circ m_{\Delta'_{1},\Delta'_{2}} \\
  & = \semg{e'} \circ m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ (id \otimes \lambda) \otimes id \circ m_{\Delta_{1},\cdot} \otimes id \circ m_{\Delta'_{1},\Delta'_{2}} \\
  & = \semg{e'} \tag{coherence}
\end{align*}

@@

Further by \cref{lem:subst},
\begin{align*}
  \semg{e'[a'/a]}
  &= \semg{e'} \circ \semg{a'/a} \tag{\cref{lem:subst}} \\
  &= \semg{e'} \tag{coherence}
\end{align*}
Again by the coherence theorem, $\semg{a'/a} = id$. Thus, $\eta$ law for $I$
holds in the denotational semantics.
\end{proof}

@@

\begin{center}
\begin{tikzcd}
  \semg{\Delta,\Delta'} \arrow[r, "m_{\Delta,\Delta'}"] &
  \semg{\Delta} \otimes \semg{\Delta'} \arrow[r, "\semg{e_1} \otimes \semg{e_2}"] &
  \semg{A} \otimes \semg{G}
\end{tikzcd}
\end{center}

@@

\begin{center}
\begin{tikzcd}
  \semg{\Delta'_1,\Delta,\Delta'_2} \arrow[r , "m_{(\Delta'_1,\Delta),\Delta'_2}"] &
  \semg{\Delta'_1,\Delta} \otimes \semg{\Delta'_2}
    \arrow[r , "m_{\Delta'_1,\Delta} \otimes id"] &
  \left(\semg{\Delta'_1} \otimes \semg{\Delta} \right) \otimes \semg{\Delta'_2}
    \arrow[d , "(id \otimes \semg{e}) \otimes id"] \\
  \semg{c} & \arrow[l , "\semg{f}"]
  \left( \left( \semg{\Delta'_1} \otimes \semg{A} \right) \otimes \semg{B}\right) \otimes \semg{\Delta'_2}
  & \left(\semg{\Delta'_1} \otimes \left( \semg{A} \otimes \semg{B}\right) \right) \otimes \semg{\Delta'_2}
    \arrow[l , "\alpha \otimes id"]
\end{tikzcd}
\end{center}

\paragraph{$\otimes\beta$}
The desired $\beta$ equality for $\otimes$ is,
\[
  \semg{\letin {(a,b)} {(a',b')} {f}} = \semg{f[a'/a,b'/b]}
\]

\begin{proof}
The left hand side reduces as follows,
\begin{align*}
  \semg{\letin {(a,b)} {(a',b')} {f}}
  & = \semg{f} \circ \alpha \otimes id \circ (id \otimes \semg{(a' , b')}) \otimes id \circ m_{\Delta,\Delta'} \\
  & = \semg{f} \circ \alpha \otimes id \circ (id \otimes \lambda \otimes \lambda \circ m_{a : A, b : B})) \otimes id \circ m_{\Delta,\Delta'} \\
  & = \semg{f} \tag{coherence}
\end{align*}

Which is equal to the right hand side,
\begin{align*}
  \semg{f[a'/a,b'/b]}
  & = \semg{f} \circ \semg{a'/a,b'/b} \\
  & = \semg{f} \tag{coherence}
\end{align*}
\end{proof}

\paragraph{$\otimes\eta$}
The desired $\eta$ equality for $\otimes$ is,
\[
  \semg{\letin {(a,b)} {c'} {f[(a,b)/c]}} = \semg{f[c'/c]}
\]

\begin{proof}
\begin{align*}
  \semg{\letin {(a,b)} {c'} {f[(a,b)/c]}}
  & = \semg{f[(a,b)/c]} \circ \alpha \otimes id \circ (id \otimes \semg{c'}) \otimes id \circ m_{\Delta,\Delta'} \\
  & = \semg{f} \circ \semg{(a,b)/c} \circ \alpha \otimes id \circ (id \otimes \semg{c'}) \otimes id \circ m_{\Delta,\Delta'} \\
  & = \semg{f} \tag{coherence}
\end{align*}

\begin{align*}
  \semg{f[c'/c]}
  &= \semg{f} \circ \semg{c'/c} \tag{\cref{lem:subst}} \\
  &= \semg{f} \tag{coherence}
\end{align*}
\end{proof}

\subsubsection{$\lto$-Functions}
% linear function
\paragraph{$\lto$-Introduction}
$\semg{\lamblto a e} : \semg{\Delta} \to \semg{A \lto B}$ is defined using the
natural isomorphism
$\phi : Hom(\semg{\Delta} \otimes \semg{A}, \semg{B}) \to Hom(\semg{\Delta}, \semg{A \lto B})$
that is provided by the adjunction between $\semg{- \otimes A}$ and $\semg{A \lto -}$.

@@

\paragraph{$\lto$-Elimination}
$\semg{\applto {e'} {e}} : \semg{\Delta,\Delta'} \to \semg{B}$ is defined by the
diagram,

\begin{center}
\begin{tikzcd}
  \semg{\Delta,\Delta'} \arrow[r , "m_{\Delta,\Delta'}"] &
  \semg{\Delta} \otimes \semg{\Delta'} \arrow[r , "id \otimes \semg{e}"] &
  \semg{\Delta} \otimes \semg{A} \arrow[r , "\phi^{-1}\left( \semg{e'} \right)"]
  &
  \semg{B}
\end{tikzcd}
\end{center}

\paragraph{$\lto\beta$}
The $\beta$ rule for $\lto$ is given by,
\[
  \semg{\applto {(\lamblto {a} {e})} {a'}} = \semg{e[a'/a]}
\]
\begin{proof}
\begin{align*}
  \semg{\applto {(\lamblto {a} {e})} {a'}}
  & = \phi^{-1}(\semg{\lamblto {a} {e}}) \circ id \otimes \semg{a'} \circ m_{\Delta,\Delta'} \\
  & = \phi^{-1}(\phi(\semg{e})) \circ id \otimes \semg{a'} \circ m_{\Delta,\Delta'} \\
  & = \semg{e} \circ id \otimes \semg{a'} \circ m_{\Delta,\Delta'} \\
  & = \semg{e} \circ id \otimes \lambda \circ m_{\Delta,\Delta'} \\
  & = \semg{e} \tag{coherence}
\end{align*}

\begin{align*}
  \semg{e[a'/a]}
  & = \semg{e} \circ \semg{a'/a} \tag{\cref{lem:subst}} \\
  & = \semg{e} \tag{coherence}
\end{align*}
\end{proof}

\paragraph{$\lto\eta$}
The $\eta$ rule for $\lto$ is given by,
\[
  \semg{\lamblto {a} {\applto e a}} = \semg{e}
\]

\begin{proof}
\begin{align*}
  \semg{\lamblto {a} {\applto e a}}
  & = \phi(\semg{\applto e a}) \\
  & = \phi(\phi^{-1}(\semg{e}) \circ id \otimes \semg{a} \circ m_{\Delta,a:A}) \\
  & = \phi(\phi^{-1}(\semg{e}) \circ id \otimes \lambda \circ m_{\Delta,a:A}) \\
  & = \phi(\phi^{-1}(\semg{e})) \tag{coherence} \\
  & = \semg{e}
\end{align*}
\end{proof}

\subsubsection{$\tol$-Functions}
\paragraph{$\tol$-Introduction}
Just as with the other linear function type, we have an adjunction between
$\semg{A \otimes -}$ and $\semg{- \tol A}$. $\semg{\lambtol a e} : \semg{\Delta} \to \semg{B \tol A}$ is defined using the
natural isomorphism
$\psi : Hom(\semg{A} \otimes \semg{\Delta}, \semg{B}) \to Hom(\semg{\Delta}, \semg{B \tol A})$
induced by this adjunction. In particular, $\semg{\lambtol a e}$ is given by $\psi$
acting on the following diagram

\begin{center}
\begin{tikzcd}
  \semg{A} \otimes \semg{\Delta} \arrow[r , "\lambda^{-1} \otimes id"] &
  \semg{a : A} \otimes \semg{\Delta} \arrow[r , "m^{-1}_{a : A , \Delta}"] &
  \semg{a : A , \Delta} \arrow[r , "\semg{e}"] &
  \semg{B}
\end{tikzcd}
\end{center}

\paragraph{$\tol$-Elimination}
The application of a $\tol$-function,
$\semg{\apptol {e} {e'}} : \semg{\Delta',\Delta} \to \semg{B}$ defined by the diagram

\begin{center}
\begin{tikzcd}
  \semg{\Delta',\Delta} \arrow[r , "m_{\Delta',\Delta}"] &
  \semg{\Delta'} \otimes \semg{\Delta} \arrow[r , "\semg{e'} \otimes id"] &
  \semg{A} \otimes \semg{\Delta} \arrow[r , "\phi^{-1}\left( \semg{e} \right)"] &
  \semg{B}
\end{tikzcd}
\end{center}

\paragraph{$\tol\beta$}

\paragraph{$\tol\eta$}
